---
title: "Replay the example datasets: from upstream output to a differential-expression report"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Replay the example datasets: from upstream output to a differential-expression report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette shows how to install `ProteoAnalystR` and **replay a full analysis
locally** on data that ships inside the package: from a raw upstream search
output, through filtering / normalization / imputation, to differential
expression and a self-documenting `analysis_summary.txt`. It uses the same
functions, in the same order, that the ProteoAnalyst web application runs; no
app, Docker, Payara, or Rserve is required.

Two small datasets are bundled, covering the two modalities and two different
upstream tools:

| dataset | modality / tool | organism | design |
|---|---|---|---|
| `fragpipe_proteomics` | protein LFQ / **FragPipe** | human | IDHmut vs IDHwt (3 vs 3) |
| `diann_phospho`       | phosphosite / **DIA-NN** | human | CTL vs DRUG (4 vs 4) |

## Requirements

- **R >= 4.5** (the manuscript environment is pinned to R 4.5.3 / Bioconductor
  3.22). The package `Imports` are Bioconductor-heavy (`limma`, `DEqMS`,
  `DESeq2`, `edgeR`, `MSstats`, `impute`, `preprocessCore`, `fgsea`, ...); the
  reproducibility deposit ships a complete `renv.lock` to recreate the exact
  environment with `renv::restore()`.
- **An organism annotation SQLite** (`hsa_genes.sqlite`) for identifier
  mapping. Point the package at the directory that contains it with the
  `OMICS_LIB_DIR` environment variable. The manuscript case-study reproduction
  bundles ship a small `db/` subset that provides exactly this file (and the
  kinase and compartment resources used below); see the *Reproducibility
  deposit* section.

## Locating the bundled data

The bundled files are read straight from the installed package, so this works
offline:

```{r locate, eval = TRUE}
library(ProteoAnalystR)

list_bundled_datasets()[, c("dataset", "role", "tool", "organism", "design")]

prot <- bundled_example("fragpipe_proteomics")
phos <- bundled_example("diann_phospho")
basename(prot)   # input + metadata
basename(phos)
```

The pipeline chunks below are shown with `eval = FALSE` so this vignette renders
anywhere. To actually run them, install the package dependencies and set
`OMICS_LIB_DIR` to a directory containing `hsa_genes.sqlite`, then paste the code
into an R session started in an empty working directory (the pipeline writes its
per-session state there).

```{r libdir, eval = FALSE}
# point identifier mapping at the annotation SQLite (adjust to your path)
Sys.setenv(OMICS_LIB_DIR = "/path/to/db")   # dir containing hsa_genes.sqlite
```

## Example 1 — FragPipe LFQ proteomics

A FragPipe `combined_protein.tsv` (MaxLFQ, protein-level) comparing IDH-mutant
vs IDH-wildtype human samples. Because this is already a protein-level matrix,
there is no peptide-to-protein summarization step.

```{r prot, eval = FALSE}
prot <- bundled_example("fragpipe_proteomics")
DATA <- prot[["input"]]; META <- prot[["metadata"]]

Init.Data(TRUE, "data/")

# FragPipe protein matrix -> MaxLFQ quant. (quantType = "protein_maxlfq")
SetFragpipeOptions("protein_maxlfq", TRUE, NA, NA)
dataSet <- ReadTabExpressData(DATA, META, "false", "default", "", "fragpipe")
ExportMetaInfoArrow(DATA)

# identifier annotation (human, protein), then QC
dataSet <- PerformDataAnnot(DATA, "hsa", "prot", "NA", "mean")
SummarizeQC(DATA, "proc", 0.1)

# FragPipe format filtering (drops contaminants / decoys / by-site rows)
ApplyFormatSpecificFiltering(DATA, "fragpipe",
  list(removeContaminants = TRUE, removeDecoys = FALSE, removeOnlyBySite = FALSE,
       minPeptides = 0, qvalueFilter = FALSE, minProb = 0.0, pepFilter = FALSE,
       pepThreshold = 0.01, spectronautQValueFilter = FALSE,
       spectronautQValueThreshold = 0.01, spectronautMinPeptides = 0))

# log2 + sample-median centering; MNAR (limit-of-detection) imputation
dataSet <- PerformNormalization(DATA, "logMedian", 15, 0, "true", "false",
                                "sum", "true", 50, "none", "__manual__")
dataSet <- ImputeMissingVar(DATA, "min")

# design + limma DE (note: the reader normalizes the group column to "Condition")
SetSelectedMetaInfo(DATA, "Condition", "NA", FALSE)
dataSet <- SetupDesignMatrix(DATA, "limma")
dataSet <- PerformDEAnal(DATA, "custom", "IDHmut vs. IDHwt", "NA", "intonly", FALSE)
dataSet <- GetSigfeatures(DATA, "Result_1", 0.05, 0.0, 1, "true")

# self-documenting parameter report
WriteAnalysisSummary("analysis_summary.txt")
```

Expected result (limma, IDHmut vs IDHwt): **4135 features tested, 1345
significant** at `adj.P.Val < 0.05` (up 625 / down 720). The
`analysis_summary.txt` Differential Expression section reads:

```
== Differential Expression ==
Method:                           limma
Comparison:                       IDHmut-IDHwt
Number of features tested:        4135
Significance:                     adj.P.Val (BH-FDR) < 0.05
Fold-change threshold (log2):     0
p-value columns in results:       P.Value (raw), adj.P.Val (BH-FDR)
Significant features:             1345  (up 625 / down 720)
```

## Example 2 — DIA-NN phosphoproteomics + enrichment

A DIA-NN phosphosite report comparing a control vs drug-treated human sample
(CTL vs DRUG). The site table carries a `SequenceWindow` column, which the motif
enrichment uses directly.

```{r phos, eval = FALSE}
phos <- bundled_example("diann_phospho")
DATA <- phos[["input"]]; META <- phos[["metadata"]]

Init.Data(TRUE, "data/")

# DIA-NN phospho: SetPhosphoOptions(format = "diann") + ReadPhosphoData(dataFormat = "diann")
SetPhosphoOptions("diann", "phospho", 0.0, c("S", "T", "Y"))
dataSet <- ReadPhosphoData(DATA, META, 0.0, "diann", "", 1.0, 1.0, FALSE, FALSE,
                           c("S", "T", "Y"))
ExportMetaInfoArrow(DATA)

dataSet <- PerformDataAnnot(DATA, "hsa", "prot", "NA", "mean")
SummarizeQC(DATA, "proc", 0.1)
dataSet <- ApplyPhosphoFormatSpecificFiltering(DATA, "diann",
  list(qvalueFilter = FALSE, minPeptides = 0))

dataSet <- PerformNormalization(DATA, "logMedian", 15, 0, "true", "false",
                                "sum", "true", 50, "none", "__manual__")
dataSet <- ImputeMissingVarPhospho(DATA, "min")   # note: phospho variant

# design + limma DE (phospho metadata keeps the "Group" column)
SetSelectedMetaInfo(DATA, "Group", "NA", FALSE)
dataSet <- SetupDesignMatrix(DATA, "limma")
dataSet <- PerformDEAnal(DATA, "custom", "CTL vs. DRUG", "NA", "intonly", FALSE)
dataSet <- GetSigfeatures(DATA, "Result_1", 0.05, 0.0, 1, "true")
dataSet <- MapPhosphositeUniprotToSymbol()

WriteAnalysisSummary("analysis_summary.txt")
```

Expected result (limma, CTL vs DRUG): **9844 sites tested, 2291 significant** at
`adj.P.Val < 0.05` (up 1191 / down 1100).

### Phosphosite enrichment

Kinase, compartment, and motif enrichment run from the DE result once
`paramSet$lib.path` points at the annotation `db/` (which holds
`ksea/NetworKIN_uniprot_hsa.qs` and `hsa/hsa_localization.qs`):

```{r phos-enrich, eval = FALSE}
paramSet <- readSet(paramSet, "paramSet")
paramSet$lib.path <- paste0(Sys.getenv("OMICS_LIB_DIR"), "/")   # dir with ksea/ and hsa/
saveSet(paramSet, "paramSet")

PerformKinaseEnrichment(DATA, "networkin")   # -> kinase_enrichment.csv
PerformCompartmentEnrichment(DATA)           # -> compartment_enrichment.csv
PerformMotifEnrichment(DATA)                 # -> motif_enrichment.csv

WriteAnalysisSummary("analysis_summary.txt") # refresh so enrichment sections appear
```

- **Kinase enrichment** runs against NetworKIN (175 kinases; top hit NEK2,
  ORA p 9.9e-6, FDR 0.0015). Needs `<lib.path>/ksea/NetworKIN_uniprot_hsa.qs`.
- **Compartment enrichment** tests broad subcellular compartments (7 compartments;
  Cytoskeleton FDR 2.1e-5 and Cell surface & adhesion FDR 0.013 are enriched).
  Needs `<lib.path>/hsa/hsa_localization.qs`; if that file is absent the call
  returns cleanly without a compartment section.
- **Motif enrichment** is self-contained (it reads the `SequenceWindow` column);
  8 broad motif classes are tested.

Each enrichment section in `analysis_summary.txt` records the number of
significant phosphosites tested, the DE-significance definition, the measured
background and its size, and the BH-FDR cutoff, for example:

```
== Kinase Enrichment ==
Number of kinases tested:         175
Significant phosphosites tested:  2291
DE significance definition:       FDR-adjusted p < 0.05
Enrichment background:            9844 quantified phosphosites
Enrichment significance cutoff:   BH-FDR < 0.05
```

## What `analysis_summary.txt` records

`WriteAnalysisSummary()` reconstructs one section per analysis that has been run
from the persisted session state, so the file always reflects what you did. For
the differential-expression and enrichment steps it states, explicitly: the
number of features / sites tested, the raw and adjusted (BH-FDR) p-value
columns, the FDR threshold, the fold-change threshold, the enrichment
multiple-testing correction, and the background/universe used for enrichment
(the measured proteome / phosphoproteome). It is offered for download from the
application's Downloads page and is written to the working directory here.


